Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-19 _tex/index.tex
Built site for gh-pages

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might examine the
question, ``To what extent do different teaching methods (\(T\))
influence students' conceptual understanding of a topic (\(Y\))?'' To
investigate this, the study could randomly assign students to two
groups, each using a different teaching method \((T_{i}=\{1,2\})\).
Students' conceptual understanding of the topic could be evaluated
through pairwise comparisons, resulting in a dichotomous outcome
\((Y_{i}=\{0,1\})\), indicating which student among those compared has a
higher level of understanding. The research question could be then
framed as the estimand, ``\emph{On average}, is there a difference in
conceptual understanding of the topic between the two groups of
students?'' This estimand could be mathematically expressed by the
associational quantity \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), where
\(E[\cdot]\) denotes the expected value. An example of this approach is
seen in \citet{Jones_et_al_2019}.

Researchers then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
numerical values that approximate the estimand and are derived through
\emph{estimation}, which refer to the process of integrating data with
an estimator \citep{Everitt_et_al_2010}. Although various methods can
approximate an estimand, researchers prioritize estimators with
desirable properties that ensure the accuracy of estimates. For
instance, the BTL model \citep{Bradley_et_al_1952, Luce_1959} is an
estimator known for its effectiveness in modeling pairwise comparisons,
yielding accurate estimates for the target of inference discussed
earlier, as long as the assumptions of the model are met.

However, many studies aim to understand the mechanisms underlying
specific data types and also seek to establish causal relationships
rather than merely infer associations. In the earlier example, the
differences between groups obtained using the BTL model, referred to as
the associational estimates, can be interpreted as causal because the
data were collected through a randomized experiment. Randomized
experiments enable the causal interpretation of associational estimates
by ensuring several key properties, including common support, no
interference, and consistency \citep{Morgan_et_al_2014, Neal_2020}. The
most crucial property, however, is that randomization effectively
eliminates confounding. \emph{Confounding} occurs when an external
variable influences both the outcome and the variable of interest,
leading to spurious associations \citep{Everitt_et_al_2010}.
Randomization mitigates this issue by decoupling the intervention
assignment mechanism, such as assigning students to different groups,
from other variables and outcomes \citep{Morgan_et_al_2014, Neal_2020}.

Experiments are widely regarded as the gold standard in evidence-based
science \citep{Hariton_et_al_2018, Hansson_2014}, however, researchers
often encounter constraints that limit their ability to conduct
experimental studies. These constraints include ethical concerns, such
as the assignment of individuals to potentially harmful interventions,
and practical limitations, such as the infeasibility of, for example,
assigning individuals to genetic modifications or physical impairments
\citep{Neal_2020}. In these instances, causal inference offers a
valuable alternative for generating causal estimates. \emph{Causal
inference} is a framework designed to identify the causes of phenomena
and estimate their effects using data
\citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical statistical
modeling, which focuses primarily on summarizing data and inferring
associations, causal inference provides a rigorous mathematical
framework for analyzing causes and counterfactuals \citep{Pearl_2009}.
Consequently, the framework offers significant theoretical insights that
enhance the design of observational and experimental studies
\citep{McElreath_2020}.

While a detailed explanation of causes and counterfactuals is beyond the
scope of this document, it is important to note that causal inference
addresses the \emph{fundamental problem of causality} through the use of
counterfactuals \citep{Neal_2020}. The framework defines
\emph{individual causal effects} (ICE) as the difference between an
individual's observed and unobserved potential outcomes. These
unobserved potential outcomes, known as \emph{counterfactuals},
represent hypothetical scenarios that are \emph{contrary to fact}, where
alternative outcomes for an individual, resulting from a specific cause,
are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The framework then extends this
estimand to \emph{average causal effects} (ACE), which captures the
difference between sample averages of observed potential outcomes and
counterfactuals. Finally, much like in an experiment, the framework
identifies the ACE from associational estimates by conditioning for a
\emph{sufficient adjustment set} of variables, ensuring the absence of
confounding \citep{Morgan_et_al_2014}.

@@

The structural approach, however, offers a notable advantage over the
potential outcomes approach by allowing the graphical and formal
representation of theories through directed acyclic graphs
\citep[DAG,][]{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
DAGs function as heuristics, effectively conveying the presumed causal
structure of a system, referred to as a \emph{scientific model}. They do
not represent detailed statistical models but allow researchers to
deduce which statistical models can provide valid causal inferences,
assuming the causal structure depicted in the DAGs are accurate
\citep{McElreath_2020}. The Identification-Estimation flowchart in
Figure~\ref{fig-IEflow} visually represents this process of
transitioning from estimands to estimates, as well as the application of
the scientific model and data to identify and estimate causal effects.

@@

The motivating example in Section~\ref{sec-framework-structural} can
further illustrate how to use the five fundamental building blocks to
construct a system's causal structure. In this scenario, the
investigator aims to determine whether, \emph{on average}, there is a
difference in conceptual understanding of a topic \((Y_{i}=\{0,1\})\)
between two groups of students \((T_{i}=\{1,2\})\), described by the
estimand \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\). However, unlike the
previous case, an experiment cannot be conducted, and the problem
suggests that the country to which a student belongs (\(X\)) may
influence both \(T\) and \(Y\). Such scenarios are plausible, especially
when the teaching methods depend on software or access to technology,
which may be limited in certain countries {(maybe an example with more
impact?)}. Figure~\ref{fig-example} illustrates the plausible causal
structure of the motivating example. A detailed examination of Figures
\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
\ref{fig-scm_example1} reveals the presence of at least four of the five
fundamental building blocks. The figures display multiple descendants,
as indicated by pairwise relations such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
features multiple pairs of unconnected nodes, evident from the relations
\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
fork \(X \rightarrow \{T,Y\}\), and two colliders with
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).

>>>>

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might examine the
question, ``To what extent do different teaching methods (\(T\))
influence students' conceptual understanding of a topic (\(Y\))?'' To
investigate this, the study could randomly assign students to two
groups, each using a different teaching method \((T=\{1,2\})\).
Students' conceptual understanding of the topic could be evaluated
through pairwise comparisons, resulting in a dichotomous outcome
\((Y=\{0,1\})\), indicating which student among those compared has a
higher level of understanding. The research question could be then
framed as the estimand, ``\emph{On average}, is there a difference in
conceptual understanding of the topic between the two groups of
students?'' This estimand could be mathematically expressed by the
associational random quantity \(E[Y| T=1] - E[Y| T=2]\), where
\(E[\cdot]\) denotes the expected value. An example of this approach is
seen in \citet{Jones_et_al_2019}.

Researchers then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
numerical values that approximate the estimand and are derived through
\emph{estimation}, which refers to the process of integrating data with
an estimator \citep{Everitt_et_al_2010}. Although various methods can
approximate an estimand, researchers prioritize estimators with
desirable properties that ensure the accuracy of estimates. For
instance, the Z-test is an estimator known for its effectiveness in
comparing two proportions, yielding accurate estimates when its
underlying assumptions are met \citep{Kanji_2006}. The Z-test is
expressed as a signal-to-noise ratio:
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is the
difference between the group sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y| T=1]\) and \(E[Y| T=2]\), respectively. The noise \(\hat{s}_{p}\)
is the unpooled sample variability observed between the two groups.

However, many studies aim to understand the mechanisms underlying
specific data and also seek to establish causal relationships rather
than merely infer associations. In the earlier example, the differences
between groups obtained using the Z-test, referred to as the
associational estimate, can be interpreted as causal because the data
were collected through a randomized experiment. Randomized experiments
enable the causal interpretation of associational estimates by ensuring
several key properties, including common support, no interference, and
consistency \citep{Morgan_et_al_2014, Neal_2020}. The most crucial
property, however, is that randomization effectively eliminates
confounding. \emph{Confounding} occurs when an external variable
influences both the outcome and the variable of interest, leading to
spurious associations \citep{Everitt_et_al_2010}. Randomization
mitigates this issue by decoupling the intervention assignment
mechanism, such as assigning students to different groups, from other
variables and outcomes \citep{Morgan_et_al_2014, Neal_2020}.

Experiments are widely recognized as the gold standard in evidence-based
science \citep{Hariton_et_al_2018, Hansson_2014}. However, researchers
often face constraints that limit their ability to conduct experimental
studies. These constraints include ethical concerns, such as the
assignment of individuals to potentially harmful interventions, and
practical limitations, such as the infeasibility of, for example,
assigning individuals to genetic modifications or physical impairments
\citep{Neal_2020}. In these situations, causal inference provides a
valuable alternative for generating causal estimates, particularly when
the goal is to understand the mechanisms underlying specific data.
Moreover, the framework offers significant theoretical insights that
enhance the design of observational and experimental studies
\citep{McElreath_2020}.

Unlike classical statistical modeling, which focuses primarily on
summarizing data and inferring associations, \emph{causal inference} is
a framework designed to identify causes and estimate their effects using
data \citep{Shaughnessy_et_al_2010, Neal_2020}. This framework employs
rigorous mathematical techniques to address the fundamental problem of
causality \citep{Pearl_2009}, which is essential for understanding and
defining causal effects. The \emph{fundamental problem of causality}
centers on answering the question, ``What would have happened `in the
world' under different circumstances?'' a concept known as
counterfactuals. \emph{Counterfactuals} represent hypothetical scenarios
that are \emph{contrary to fact}, where alternative outcomes resulting
from a specific cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}.

Although a comprehensive discussion of causes and counterfactuals
exceeds the scope of this document, a brief overview of how the
framework addresses the fundamental problem of causality is possible
using the previous example. The framework starts by defining
\emph{individual causal effects} (ICE) as the difference between
students' observed and unobserved potential outcomes:
\(\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)\). Notice once a
student is assigned to \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) is no longer observed nor observable, and is thus termed a
\emph{counterfactual}. To address the challenge posed by
counterfactuals, the framework extends the ICE to \emph{average causal
effects} (ACE). The ACE is defined as
\(\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), and
represents the difference between the average of observed potential
outcomes and counterfactuals in the sample. Finally, akin to
experimental studies, the ACE is identified from associational estimates
by ensuring the absence of confounding. This is accomplished by
statistically conditioning on a \emph{sufficient adjustment set} of
variables (\(X\)). Consequently, the ACE is expressed as
\(\tau = E_{X}[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]]\), where
\(E_{X}[\cdot]\) denotes the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}.

@@

The structural approach, however, offers a notable advantage over the
potential outcomes approach by allowing the graphical and formal
representation of the ACE through directed acyclic graphs
\citep[DAG,][]{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
DAGs function as heuristics, effectively conveying the presumed causal
structure of a system, referred to as a \emph{scientific model}. They do
not represent detailed statistical models but allow researchers to
deduce which statistical models can provide valid causal inferences,
assuming the causal structure depicted in the DAGs are accurate
\citep{McElreath_2020}. The Identification-Estimation flowchart in
Figure~\ref{fig-IEflow} visually represents the process of transitioning
from estimands to estimates, as well as the application of the
scientific model and data to identify and estimate causal effects.

@@

The motivating example in Section~\ref{sec-framework-structural} can
further illustrate how to use the five fundamental building blocks to
construct a system's causal structure. In this scenario, the
investigator aims to determine whether, \emph{on average}, there is a
difference in conceptual understanding of a topic \((Y=\{0,1\})\)
between two groups of students \((T=\{1,2\})\), described by the
estimand \(E[Y| T=1] - E[Y| T=2]\). However, unlike the previous case,
an experiment cannot be conducted, and the problem suggests that the
country to which a student belongs (\(X\)) may influence both \(T\) and
\(Y\). Such scenarios are plausible, especially when the teaching
methods depend on software or access to technology, which may be limited
in certain countries {(maybe an example with more impact?)}.
Figure~\ref{fig-example} illustrates the plausible causal structure of
the motivating example. A detailed examination of Figures
\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
\ref{fig-scm_example1} reveals the presence of at least four of the five
fundamental building blocks. The figures display multiple descendants,
as indicated by pairwise relations such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
features multiple pairs of unconnected nodes, evident from the relations
\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
fork \(X \rightarrow \{T,Y\}\), and two colliders with
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).
